---
title: "class12"
author: "Abraham Rachlin"
format: pdf
---

## Section 4: Population Scale Analysis

How many samples are there?

>Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```
```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

```{r}
library(ggplot2)
```

The median for A/A appears to be 31-32, the median for A/G appears to be 25, and the median for G/G appears to be around 20.

>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

What you could infer about the expression value between A/A and G/G in the boxplot below is that A/A's expression valyue has a mean that is almost 10 values higher than that of G/G. G/G's maximum is not even A/A's minimum. Having a G/G does appeat to affect the expression of ORMDL3, as it is much lower than the others, therefore affecting its expression of the gene.

```{r}
ggplot(expr, aes(geno, exp, fill=geno)) +
  geom_boxplot()
```


