---
title: "class15 mini project"
author: "Abraham Rachlin"
format: pdf
---

```{r}
library(DESeq2)
```

```{r}
metaFile <- "GSE37704_metadata.csv"
countFile <- "GSE37704_featurecounts.csv"

# Import metadata and take a peak
colData = read.csv(metaFile, row.names=1)
head(colData)
# Import countdata
countData = read.csv(countFile, row.names=1)
head(countData)
```
>Q. Complete the code below to remove the troublesome first column from countData

```{r}
countData <- as.matrix(countData[,-1])
head(countData)
```

>Q. Complete the code below to filter countData to exclude genes (i.e. rows) where we have 0 read count across all samples (i.e. columns).

```{r}
to.keep <- rowSums(countData[,1:2] == 0) == 0
sum(to.keep)

countData = countData[to.keep, ]
head(countData)
```

```{r}
dds = DESeqDataSetFromMatrix(countData = countData,
                             colData=colData,
                             design=~condition)
dds = DESeq(dds)
```
```{r}
dds
```
```{r}
res = results(dds, contrast=c("condition", "hoxa1_kd", "control_sirna"))
```

>Q. Call the summary() function on your results to get a sense of how many genes are up or down-regulated at the default 0.1 p-value cutoff.

```{r}
summary(res)
```

```{r}
plot( res$log2FoldChange, -log(res$padj) )
```

>Q. Improve this plot by completing the below code, which adds color and axis labels

```{r}
# Make a color vector for all genes
mycols <- rep("gray", nrow(res) )

# Color red the genes with absolute fold change above 2
mycols[ abs(res$log2FoldChange) > 2 ] <- "red"

# Color blue those with adjusted p-value less than 0.01
#  and absolute fold change more than 2
inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

plot( res$log2FoldChange, -log(res$padj), col= mycols, xlab="Log2(FoldChange)", ylab="-Log(P-value)" )
```

>Q. Use the mapIDs() function multiple times to add SYMBOL, ENTREZID and GENENAME annotation to our results by completing the code below.

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
```
```{r}
columns(org.Hs.eg.db)

res$symbol = mapIds(org.Hs.eg.db,
                    keys=row.names(res), 
                    keytype="ENSEMBL",
                    column="SYMBOL",
                    multiVals="first")

res$entrez = mapIds(org.Hs.eg.db,
                    keys = row.names(res),
                    keytype = "ENSEMBL",
                    column = "ENTREZID",
                    multiVals = "first")

res$name = mapIds(org.Hs.eg.db,
                  keys = row.names(res),
                  keytype = "ENSEMBL",
                  column = "GENENAME",
                  multiVals = "first")

head(res, 10)

```

>Q. Finally for this section let's reorder these results by adjusted p-value and save them to a CSV file in your current project directory.

```{r}

res <- res[order(res$padj), ]
write.csv(res, file = "deseq_results.csv")

```

```{r}
library(pathview)
```
```{r}
library(gage)
library(gageData)

data(kegg.sets.hs)
data(sigmet.idx.hs)
```
```{r}
# Focus on signaling and metabolic pathways only
kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]

# Examine the first 3 pathways
head(kegg.sets.hs, 3)
```
```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```


```{r}
keggres = gage(foldchanges, gsets=kegg.sets.hs)
```

```{r}
attributes(keggres)
```
```{r}
head(keggres$less)
```
```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110")
```

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110", kegg.native=FALSE)
```

```{r}
keggrespathways <- rownames(keggres$greater)[1:5]
```

```{r}
keggresids = substr(keggrespathways, start=1, stop=8)
keggresids
```
```{r}
pathview(gene.data=foldchanges, pathway.id=keggresids, species="hsa")
```

>Q. Can you do the same procedure as above to plot the pathview figures for the top 5 down-reguled pathways?

```{r}

keggrespathways_down <- rownames(keggres$less)[1:5]  
keggresids_down <- substr(keggrespathways_down, start = 1, stop = 8)

print(keggresids_down)
```

```{r}
data(go.sets.hs)
data(go.subs.hs)
```
```{r}
gobpsets = go.sets.hs[go.subs.hs$BP]

gobpres = gage(foldchanges, gsets=gobpsets, same.dir=TRUE)

lapply(gobpres, head)
```
```{r}
sig_genes <- res[res$padj <= 0.05 & !is.na(res$padj), "symbol"]
print(paste("Total number of significant genes:", length(sig_genes)))

```

```{r}
write.table(sig_genes, file="significant_genes.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
```

>Q: What pathway has the most significant “Entities p-value”? Do the most significant pathways listed match your previous KEGG results? What factors could cause differences between the two methods?

The pathway that is most significant is cell cycle. The most significant does not match, as it is not the same pathway. The factors that could have affected it is potential error in the code on my end.